---
title: 'Analysis of droughts caused by climate variability at Lake Poopó, Bolivia'
author: "Magdalena Halbgewachs"
date: "29 April 2019"
output:
  ioslides_presentation: default
  beamer_presentation: default
---
```{r, include=FALSE}
#slides > slide.backdrop {
 #background: white;
#}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##investigation area Lake Poopó
- located in the Bolivian Altiplano
- mean elevation: 3810 m a.s.l.
- Landsat time series from 1989 to 2018

```{r, echo=FALSE, warning=FALSE, error=FALSE}
library(maps)
library(mapdata)
map <- map('worldHires','Bolivia', col = 'grey', fill=T)
points(-67.056825,-18.735300,col= 'blue',pch=16, size=2)
map('lakes', add=T, fill=T, col = 'blue', boundary='blue')
plot(map)
```


##investigation area Lake Poopó
- extremely regulated by environmental influences
- declared dry in 2015 but still fills periodically with water
- Desaguadero River: transports water from Lake Titicaca to Lake Poopó

```{r, echo=FALSE, warning=FALSE, error=FALSE}
library(maps)
library(mapdata)
map <- map('worldHires','Bolivia', col = 'grey', fill=T)
points(-67.056825,-18.735300,col= 'blue',pch=16, size=2)
map('lakes', add=T, fill=T, col = 'blue', boundary='blue')
plot(map)
```


##NDWI
NDWI=(GREEN-NIR)/(GREEN+NIR)

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, results='hide'}
library(rgdal)
library(raster)
library(ggplot2)
library(rasterVis)
library(maptools)

NDWI <- "D:/01_Uni/02_Master/MB1_Digital Image Analysis and GIS/00_final_project/01_Landsat/NDWI"
all_NDWI <- list.files(NDWI, full.names = TRUE, pattern = ".tif$")
#----------------------------------------------------------------------------------------
# Create a raster stack of the NDWI time series
NDWI_stack <- stack(all_NDWI)
# gsub to modify label names
rasterNames  <- gsub("NDWI_","", names(NDWI_stack))
#----------------------------------------------------------------------------------------
#levelplot
cols <- colorRampPalette(brewer.pal(11,"Spectral"))    #  number of colors

ndwi <- levelplot(NDWI_stack,
          main="NDWI\nLake Poopó, Bolivia (1989 - 2018)",
          col.regions=cols,
          names.attr=rasterNames,                     # using new defined names
          scales=list(draw=FALSE ))                  # remove axes labels & ticks
plot(ndwi)
```


## lake area extent and bathymetry
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, results='hide'}
class <- "D:/01_Uni/02_Master/MB1_Digital Image Analysis and GIS/00_final_project/01_Landsat/classification"
all_class <- list.files(class, full.names = TRUE, pattern = ".tif$")
#----------------------------------------------------------------------------------------
# Create a raster stack of the NDWI time series
class_stack <- stack(all_class)
# gsub to modify label names
rasterNames  <- gsub("class_","", names(class_stack))
#----------------------------------------------------------------------------------------
#load bathymetric lines
bath.shp <- readOGR("D:\\01_Uni\\02_Master\\MB1_Digital Image Analysis and GIS\\00_final_project\\Lake_Poopo\\bathymetry_qgis.shp")
#----------------------------------------------------------------------------------------
#levelplot
cols <- rasterTheme(region = brewer.pal("Blues", n=3))
cuts <- c(0.1, 0.5, 1.1)

classif <- levelplot(class_stack,
          main="Lake area\nLake Poopó, Bolivia (1989 - 2018)",
          par.settings=cols,
          at=cuts,
          names.attr=rasterNames,
          kolorkey=F,
          scales=list(draw=FALSE ))+
  layer(sp.polygons(bath.shp), packets = (1:22))
plot(classif)
```


##change detection
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, , results='hide'}
minus <- "D:/01_Uni/02_Master/MB1_Digital Image Analysis and GIS/00_final_project/01_Landsat/change_minus"
all_minus <- list.files(minus, full.names = TRUE, pattern = ".tif$")
#----------------------------------------------------------------------------------------
# Create a raster stack of the NDWI time series
minus_stack <- stack(all_minus)
# gsub to modify label names
rasterNames  <- gsub("change_","", names(minus_stack))

#----------------------------------------------------------------------------------------
#levelplot
cols <- colorRampPalette(brewer.pal(9,"RdBu"))

change_det <- levelplot(minus_stack,
          main="Loss and gain of water area\nLake Poopó, Bolivia (1989 - 2018)",
          col.regions=cols,
          names.attr=rasterNames,
          scales=list(draw=FALSE ))
plot(change_det)
```


```{r, warning=FALSE}
#loading required packages
library(ggplot2)
library(patchwork)
library(reshape2)
library(corrplot)

##########################################################################################
# data import
##########################################################################################
###precipitation and evapotranspiration all months 1989 - 2018###
df_data_all <- read.csv("C:\\02_Studium\\02_Master\\01_Semester 1\\00_paper_work\\01_Lakes\\Lake_Poopó\\prec_et_mean_all.csv", header=T, sep=";")
if(names(df_data_all)[1]=="ï..YEAR"){             #replaces mysterious column name
  names(df_data_all)[1]<- "YEAR"
}
df_data_all[[1]] <- as.Date(df_data_all[[1]], origin="1989-01.01")   #read as date
names(df_data_all) <- c ("YEAR", "PREC", "ET")  #rename columns

#----------------------------------------------------------------------------------------
###precipitation and evapotranspiration April and July 1989 - 2018###
df_data <- read.csv("C:\\02_Studium\\02_Master\\01_Semester 1\\00_paper_work\\01_Lakes\\Lake_Poopó\\prec_et_april_july.csv", header=T, sep=";")
if(names(df_data)[1]=="ï..YEAR"){             #replaces mysterious column name
  names(df_data)[1]<- "YEAR"
}
df_data[[1]] <- as.Date(df_data[[1]], origin="01-04-1989")   #read as date
names(df_data) <- c ("YEAR", "PREC_APRIL", "PREC_JULY", "ET_APRIL", "ET_JULY")

#----------------------------------------------------------------------------------------
###lake area###
##only if df is not in current environment

area_april<- read.csv("D:\\01_Uni\\02_Master\\MB1_Digital Image Analysis and GIS\\00_final_project\\01_Landsat\\area_april.csv", header=T, sep=",")
area_april <- area_april[-c(1)] #remove first column
if(names(area_april)[1]=="ï..YEAR"){
  names(area_april)[1]<- "YEAR"
}
area_april[[1]] <- as.Date(area_april[[1]], origin="1989-04-1")

area_july <- read.csv("D:\\01_Uni\\02_Master\\MB1_Digital Image Analysis and GIS\\00_final_project\\01_Landsat\\area_july.csv", header=T, sep=",")
area_july <- area_july[-c(1)] #remove first column
if(names(area_july)[1]=="ï..YEAR"){
  names(area_july)[1]<- "YEAR"
}
area_july[[1]] <- as.Date(area_july[[1]], origin="1989-07-1")

#----------------------------------------------------------------------------------------
###save all data from April and July in one dataframe###
df_data <- merge (df_data, area_april,by="YEAR", all=T)
df_data <- merge (df_data, area_july, by="YEAR", all=T)
names(df_data) <- c ("YEAR", "PREC_APRIL", "PREC_JULY", "ET_APRIL", "ET_JULY", "A_APRIL", "A_JULY")


#----------------------------------------------------------------------------------------
###lake area Titicaca###

titicaca <- read.table("D:\\01_Uni\\02_Master\\MB1_Digital Image Analysis and GIS\\00_final_project\\Titicaca\\time_series_titicaca", sep = "", header = T, fill = T)
if(names(titicaca)[1]=="ï..YEAR"){
  names(titicaca)[1]<- "YEAR"
}
titicaca[[1]] <- as.Date(titicaca[[1]])
```


##lake area extent in km²

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height = 5, fig.width =8}
library(ggplot2)
library(patchwork)
library(reshape2)
#library(corrplot)
#water area in April and July 1989 to 2019
water_plot <- ggplot(data=df_data, aes(x=YEAR, y=A_APRIL, group=1, color=legend)) +
  #data
  geom_bar(data=df_data, aes(x=YEAR, y=A_APRIL, color="April"), stat="identity", alpha=0.2, width=300, fill="transparent", size=1)+
  geom_bar(data=df_data, aes(x=YEAR, y=A_JULY, color="July"), stat="identity", alpha=0.2, width=300, fill="transparent", size=1)+
  #linear regression
  geom_smooth(data=df_data, aes(x=YEAR, y=A_APRIL), method='lm',formula=y~x, color="#F8766D", linetype="dashed", se=F)+
  geom_smooth(data=df_data, aes(x=YEAR, y=A_JULY), method='lm',formula=y~x, color="#00BFC4", linetype="dashed", se=F)+
  #themes
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 6),
        axis.text.y=element_text(angle = 45, size = 9))+
  #axis
  xlab("year")+
  ylab("area (km²)")+
  scale_x_date(breaks = as.Date(c("1989-04-01", "1995-04-01", "1999-04-01",
                                  "2005-04-01", "2009-04-01", "2013-04-01",
                                  "2015-04-01", "2018-04-01")),date_labels = "%Y")+
  #ggtitle("Water area of Lake Poopó in April and July (1989-2018)")+
  labs(caption = "data: based on NDWI calculations (Landsat 5 and 8)")+
  theme_gray(base_size = 15)
plot(water_plot)
```


##precipitation

```{r, echo=FALSE, warning=FALSE,  fig.height = 5, fig.width =8}
#precipitation only April and July 1989 to 2019
prec_plot <- ggplot(data=df_data, aes(x=YEAR, y=PREC_APRIL, group=1, color=legend)) +
  #data
  geom_bar(data=df_data, aes(x=YEAR, y=PREC_APRIL, color="April"),  stat="identity", alpha=0.3, width=300, fill="transparent", size=1) +
  geom_bar(data=df_data, aes(x=YEAR, y=PREC_JULY, color="July"),  stat="identity", alpha=0.3, width=300, fill="transparent", size=1)+
  #linear regression
  geom_smooth(aes(x=YEAR, y=PREC_APRIL), formula=y~x, method='lm', color="#F8766D", linetype="dashed",  se=F)+
  geom_smooth(aes(x=YEAR, y=PREC_JULY), formula=y~x, method='lm', color="#00BFC4", linetype="dashed", se=F)+
  #themes
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 6),
        axis.text.y=element_text(angle = 45, size = 9))+
  #axis
  xlab("year")+
  ylab("precipitation (mm)")+
  scale_x_date(breaks = as.Date(c("1989-04-01", "1995-04-01", "1999-04-01",
                                  "2005-04-01", "2009-04-01", "2013-04-01",
                                  "2015-04-01", "2018-04-01")),date_labels = "%Y")+
  #ggtitle("Mean precipitation in the Lake Poopó area in April and July (1989-2018)")+
  labs(caption = "data source: Servicio Nacional de Meteorología e Hidrología de Bolivia")+
  theme_gray(base_size = 15)
plot(prec_plot)
#plot(prec_plot+water_plot)
```


##evapotranspiration

```{r, echo=FALSE, warning=FALSE,  fig.height = 5, fig.width =8}
et_plot <- ggplot(data=df_data, aes(x=YEAR, y=ET_APRIL, group=1, color=legend)) +
  #data
  geom_bar(data=df_data, aes(x=YEAR, y=ET_APRIL, color="April"), stat="identity", alpha=0.3, width=300, fill="transparent", size=1)+
  geom_bar(data=df_data, aes(x=YEAR, y=ET_JULY, color="July"), stat="identity", alpha=0.3, width=300, fill="transparent", size=1)+
  #linear regression
  geom_smooth(data=df_data, aes(x=YEAR, y=ET_APRIL), method='lm',formula=y~x, color="#F8766D", linetype="dashed", se=F)+
  geom_smooth(data=df_data, aes(x=YEAR, y=ET_JULY), method='lm',formula=y~x, color="#00BFC4", linetype="dashed", se=F)+
  #themes
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 6),
        axis.text.y=element_text(angle = 45, size = 9))+ 
  #axis
  xlab("year")+
  ylab("evapotranspiration (mm)")+
  scale_x_date(breaks = as.Date(c("1989-04-01", "1995-04-01", "1999-04-01",
                                  "2005-04-01", "2009-04-01", "2013-04-01",
                                  "2015-04-01", "2018-04-01")),date_labels = "%Y")+
  #ggtitle("Mean evapotranspiration in the Lake Poopó area in April and July (1989-2018)")+
  labs(caption = "data source: Servicio Nacional de Meteorología e Hidrología de Bolivia")+
  theme_gray(base_size = 15)
plot(et_plot)
#plot(et_plot+water_plot)
```


##Lake Titicaca water levels

```{r,  echo=FALSE, warning=FALSE,  fig.height = 5, fig.width =8}
titicaca <- read.table("D:\\01_Uni\\02_Master\\MB1_Digital Image Analysis and GIS\\00_final_project\\Titicaca\\time_series_titicaca", sep = "", header = T, fill = T)
if(names(titicaca)[1]=="ï..YEAR"){
  names(titicaca)[1]<- "YEAR"
}
titicaca[[1]] <- as.Date(titicaca[[1]])

water_titicaca <- ggplot(data=titicaca, aes(x=date, y=heights, group=1))+
  #vizualisation
  geom_line(color="green4")+
  geom_point(size=1)+
  geom_smooth(method='lm',formula=y~x, se=F, color="green4")+
  #themes
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))+
  theme(legend.title = element_text(size = 10, face = "bold"))+
  theme(legend.text = element_text(size = 6))+
  theme(axis.text.y=element_text(angle = 90, hjust=1, size = 9))+
  #axis
  xlab("year")+
  ylab("water level (m a.s.l.)")+
  labs(caption = "data: Schwatke et al. 2015)+
  theme_gray(base_size = 15)
plot(water_titicaca)
```



##correlations

```{r, echo=FALSE,  fig.height = 5, fig.width =7}
#loading data
corr <- read.csv("C:\\02_Studium\\02_Master\\01_Semester 1\\00_paper_work\\01_Lakes\\Lake_Poopó\\correlation.csv", header=T, sep=";")
#rename columns
colnames(corr)[1:4] <- c("evapotranspiration", "precipitation", "area Poopó", "area Titicaca") # handling missing values
#matrix
corr_na <-cor(corr, use = "complete.obs")
res <- cor(corr_na)
#round(res, 2)

#plot correlations
corrplot <- corrplot(res, method="number", tl.col = "black", type="upper", order="hclust", tl.srt = 45)
plot(corrplot)
```


##Thank you for your attention!